---
title: "COVID-19 cases and deaths"
author: "Attila Szuts"
date: "28/11/2020"
output:
  prettydoc::html_pretty:
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
library(tidyverse); library(estimatr); library(scales); library(lspline); library(texreg); library(car)
# read in data 
df <- read_csv('data/clean/covid_pop_clean.csv')
```

# Executive summary

In this report I am going to summarise my main findings on the pattern of association between confirmed COVID-19 cases and deaths on a given day on a country level. I used a simple linear regression with log transformed predictor and outcome variables. I found a clear linear relationship between the log of number of confirmed cases and log of number of deaths: countries that have one percent higher confirmed cases, have roughly one percent higher deaths, on average $(\beta=1.031)$; however causal relationship has not been investigated in any way. The selected model can explain 89% of the variance. $(R^2=0.8859)$ This suggests that in general as the number of cases increase the number of deaths increase in the same pace, i.e. very broadly speaking high and low number of confirmed cases have the same fraction of deaths. Adding further predictor variables (like population density and response to pandemic situation) could help us better predic the number of cases and explain outliers better. 

# Introduction

My main variables are countries, confirmed cases, deaths and total population of a country. Numbers on confirmed cases and deaths were collected by the Center for Systems Science and Engineering at Johns Hopkins University. Population data for each country was downloaded from the World Bank's site. 
The population of my analysis are the world's countries during the pandemic and I am analysing a cross-sectional subset on 2020-09-10. Data quality can vary from country to country depending on the countries' healthcare system's level of sophistication.

# Histogram and summary statistics

In the histograms below we can see all of our variables are skewed to the left, having a long right tail. These extreme values are not measurement errors, but are in fact countries with extremely high numbers of COVID-19 cases. These countries include Brazil, India, and the United states with the top 3 number of confirmed cases; Brazil, India and Mexico with the top 3 number of deaths. These extreme values could be due to a number of factors, including higher than average population, population density and lack of counter-pandemic measures.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# check histograms and extreme values
df %>%
  select(confirmed, death, population) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = "free", nrow = 1) +
  geom_histogram() +
  labs(x = 'Variable values', y = 'Absolute Frequency', title = 'Histogram of numeric variables in dataset') + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) 
```


# Variable transformations

Since we would like to model percentage differences in this analysis (for countries that have higher number of cases, how will the average number of deaths change), it would make sense to transform our $x$ and $y$ variables logarithmically (if it also makes sense from a statistical point of view). To investigate this question, I plotted my data on a scatterplot and used the previous histograms and summary statistics to establish, that indeed the clearest linear pattern of association emerges when both the predictor and outcome variable is transformed. Also, my data is cross-sectional and taking logs of variables we can solve the lack of baseline for comparison. 

Because of log transformations, I had to exclude countries, where the number of confirmed cases or deaths were 0. There were no countries that did not have at least a single case, but there were 11 countries for example Mongoloia, Bhutan and Cambodia that all had 0 deaths.

```{r include=FALSE}
# excluding cases where there are no deaths, as the log of 0 is not interpretable.
df <- df %>% 
  filter(death != 0 & confirmed != 0)
# create log transformed variables
df <- df %>% 
  mutate(ln_cases = log(confirmed),
         ln_death = log(death),
         ln_cases_sq = ln_cases ^ 2)
```

# Model of choice: simple linear regression

In my analysis I used different models to try to uncover the pattern of association between confirmed cases and deaths and to try to find the model that can explain the most variance in my data while also being easily interpreted. Because of these reasons I chose the simple linear regression that I am going to introduce in the following section. The formula for this model is: 

$$ln(\text{deaths}) = \alpha + \beta \times ln(\text{cases})$$
$$\alpha = -4.272$$
$$\beta = 1.031$$
These parameters can be interpreted as the number of deaths is higher 1.031% on average for observations with 1% higher confirmed cases. The intercept ($\alpha$) means, that when there is only one confirmed case the average of the log of deaths is -4.272 (which is meaningless in this context).

```{r message=FALSE, warning=FALSE, include=FALSE}

# reg1 - simple linear regression
reg1 <- lm_robust(ln_death ~ ln_cases, data = df)

# reg2 - quadratic linear regression
reg2 <- lm_robust(ln_death ~ ln_cases + ln_cases_sq, data = df)

# reg3 - PLS
cutoff <- quantile(df$confirmed, .90)
ln_cutoff <- log(cutoff)
reg3 <- lm_robust(ln_death ~ lspline(ln_cases, ln_cutoff), data = df)

# reg4 - weighted OLS
reg4 <- lm_robust(ln_death ~ ln_cases, data = df, weights = population)
```

```{r echo=FALSE, results='asis'}
# model comparison table
htmlreg(list(reg1 , reg2 , reg3 , reg4),
        type = 'html',
        custom.model.names = c("Simple linear","Quadratic linear","PLS", "Weighted OLS"),
        caption = "Modelling case fatality and confirmed COVID-19 cases",
       custom.coef.names = c('Intercept', 'ln(cases)', 'ln(cases)^2', 'ln(cases<339062.1)', 'ln(cases>=339062.1)'),
        include.ci = FALSE,
       fontsize = 'small')
```


# Hypothesis testing and residual analysis

Next, I would like to investigate, if the true $\beta$ is 0 or not, or said otherwise: if there is a true relationship between our $x$ and $y$. So for this I set up the following hypothesis test:
$$H_0 : \beta_{true}=0, \space H_A : \beta_{true}\neq 0$$
```{r include=FALSE}
## hypothesis testing
# hypothesis testing
linearHypothesis( reg1 , "ln_cases = 0")

## residual analysis
# Get the predicted y values from the model
df$reg1_y_pred <- reg1$fitted.values
# Calculate the errors of the model
df$reg1_res <- df$ln_death - df$reg1_y_pred 
# calculate the predicted number of deaths (inverse log transformation)
df$reg1_y_pred_exp <- exp(df$reg1_y_pred)

# Find countries with largest negative errors
l1 <- df %>% top_n( -1 , reg1_res ) %>% 
  select( country , confirmed, death , reg1_y_pred_exp, reg1_y_pred , ln_death , reg1_res ) %>% 
  arrange(reg1_res)

# Find countries with largest positive errors
u1 <- df %>% top_n( 1 , reg1_res ) %>% 
  select( country , confirmed, death , reg1_y_pred_exp, reg1_y_pred , ln_death , reg1_res ) %>% 
  arrange(desc(reg1_res))
```

The estimated t-statistics is `r round(reg1$statistic[2],2)`, with p-value: `r format(reg1$p.value[2], digits = 3, scientific = 5)`, which means our test is significant at 1%. Thus I reject the $H_0$, which means the confirmed cases is not uncorrelated with the number of deaths.

Next, I investigated the residuals: 

The largest negative deviance from the predicted value is found in `r l1[[1,1]]` with predicted number of deaths of `r round(l1[[1,4]],0)`, but the real value is only `r round(l1[[1,3]],0)`. This means, that in Singapore the situation is better, than the model would predict based on the number of active cases.
The largest positive deviance from the predicted value is found in `r u1[[1,1]]` with predicted number of deaths `r round(u1[[1,4]],0)`, but the real value is `r round(u1[[1,3]],0)`. This means, that in Yemen, the situation is worse, than the model would predict based on the number of active cases.

# Appendix

## Transforming variables

```{r echo=FALSE, message=FALSE, warning=FALSE}

# loess 
df %>% 
  ggplot(aes(confirmed, death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(method = "loess", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases', 
       y = 'Number of COVID-19 deaths',
       title = 'COVID-19 cases and deaths') +
  theme_bw()

df %>% 
  ggplot(aes(confirmed, death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(method = "loess", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths',
       title = 'COVID-19 cases and deaths') +
  scale_x_continuous( trans = log_trans(), breaks = c(1,2,5,10,20,50,100,200,500,1000,10000)) + 
  theme_bw() 

df %>% 
  ggplot(aes(confirmed, death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(method = "loess", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths') +
  scale_y_continuous( trans = log_trans(), breaks = c(1,2,5,10,20,50,100,200,500,1000,10000)) + 
  theme_bw() 

df %>% 
  ggplot(aes(confirmed, death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(method = "loess", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths') +
  scale_x_continuous( trans = log_trans(), breaks = c(1,2,5,10,20,50,100,200,500,1000,10000)) + 
  scale_y_continuous( trans = log_trans(), breaks = c(1,2,5,10,20,50,100,200,500,1000,10000)) + 
  theme_bw()
```


## Estimating different models

```{r echo=FALSE, message=FALSE, warning=FALSE}
# reg1 - simple linear regression
df %>% 
  ggplot(aes(ln_cases, ln_death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(method = "lm", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths',
       subtitle = 'Simple linear regression') +
  theme_bw() 


# reg2 - quadratic linear regression
df %>% 
  ggplot(aes(ln_cases, ln_death)) + 
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) + 
  geom_smooth(method = "lm", formula = y ~ poly(x,2), color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths',
       subtitle = 'Quadratic linear regression') +
  theme_bw() 

# reg3 - PLS
df %>% 
  ggplot(aes(ln_cases, ln_death)) +
  geom_point(color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(formula = y ~ lspline(x, ln_cutoff), method = "lm", color = 'green') +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths',
       subtitle = 'PLS') +
  geom_vline(xintercept = ln_cutoff, linetype = 'dotted') + 
  theme_bw() 

# reg4 - weighted OLS
df %>% 
  ggplot(aes(ln_cases, ln_death)) + 
  geom_point(aes(size = population), color = 'purple', shape = 16, show.legend = F, alpha = 0.6) +
  geom_smooth(aes(weight = population), method = "lm", formula = y ~ x, color = 'green') +
  scale_size(range = c(1, 15)) +
  labs(x = 'Number of confirmed COVID-19 cases, log transformed', 
       y = 'Number of COVID-19 deaths, log transformed',
       title = 'COVID-19 cases and deaths',
       subtitle = 'Weighted OLS') +
  theme_bw() 
```

